---
title: "structural_model"
author: "Laura"
date: "4/21/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

The structural model:

- describes relationship between latent variables 
- order and sequence of constructs: based on theory/ experiences and knowledge
- path-models from left to right
-- left: independent variables
-- right: dependent variables
-- left: sequential ahead, influence variables on the right


## Slide with Bullets

The structural model 2 - exogeneous vs. endogeneous variables:

- variables can serve as dependent and independent
- latent variables serving only as independent = exogeneous latent variables (Y1 and Y2)
- latent variables serving only as dependent (Y4) or as independent and dependent (Y3) = endogeneous latent variables (Y3)
- each variable, which has only arrows coming from it = exogeneous latent variable
- endogeneous latent variables can have either incoming and outgoing arrows (Y3) or only incoming arrows (Y4)
- exogeneous latent variables have no error terms (because these constructs depict the independent variables explaining the dependent variables in the path-model)

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## What is the structural model?

```{r setup}


```

