---
title: "Preparing your Data for SEM Estimation"
subtitle: "Basic Cleaning and Renaming for SEMinRc"
author: "Wir"
#institute: "RStudio, PBC"
date: "2021/04/21 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    #lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(seminr)
library(knitr)
```
# Data preparation

1. Data requirements
2. Recoding variables 
3. Renaming variables

???
In this video, I will outline the basics of data preparation for SEMinR.
This means making sure that SEMinR is able to correctly interpret your data and estimate a model.
First, I will give an overview of requirements for data to be used in SEMinR estimations.
Then I will talk in detail about recoding and renaming of data. 

---

# Requirements

Or why do I even need to do anything to my data?

* data types
* variable direction
* variable range
* variable names
* na values
* ensure data parametricity (distribution in a variable)
* remove low quality responses
* treat outliers
* treat missing data



???
Before you can start with model estimation, you have to prepare your data.
As you can see, there is a plethora of things you can do to your data before writing even one line of SEMinR code.
However, in this video, I will focus on basic preparation steps.
This means making sure your data can be used by the algorithm to estimate a model.
I will demonstrate each step using R.
If you are used to do these things in a software like SPSS, you are of course free to do so.

---

# What I am talking about

```{r, echo = FALSE, include = TRUE}
mobi[1:4,1:6] %>%
  rename_with(~ c(
    "Quality Expectation",
    "Expectation Products",
    "Problem Expectation",
    "Satisfaction Overall",
    "Expectation Fulfillment",
    "Ideal Provider"
    )) %>%
knitr::kable(format = 'html',
             row.names = TRUE)
  
```


???
If you are an experienced researcher, you will be familiar with many of the data preparation steps detailed on the previous stage.
We will also share a similar language when talking about a data,
But just to make sure there are no misunderstandings, I want to define some terms for this video.
Let's assume we start out with a bunch of data collected in a survey.
As you can see, here we have a small slice of such a data set with
* 6 variables, i.e. columns, and
* 4 observations, i.e., rows

---

# Recoding variables

* rename variables
* imputate missing observations or get rid of variables


???
SEM only uses interval scaled data.
Answers should therefore be collected using measures like Likert scales or percentages of agreement.
Categorical data, e.g. gender or profession, cannot be used in SEM constructs.
The answers data should not only be numerical, but also, the numbers should represent points on a equidistant scale.
If you have reverse-coded data, you should invert them.
SEMinR also cannot handle missing data.
This means if there is only one NA in a variable used in the SEM, the whole observation is discarded.
Therefore, you should carefully consider what to do with NA values.
If not that many observations contain missing data for a certain variable, you might imputate data.
Otherwise, it might make more sense to not use the variable in your SEM.

---

# Renaming variables

???
As you can see, I have very descriptive variable names in my dataset.
If you use SEMinR, you should probably rename thos variables.
Not only does it make for a more clear path diagram.
It also speeds up model specification.
In SEMinR, you can use the multi_items() function you describe your measurement model.
For this, SEMinR assumes all your variables are named after the patter prefix-number.
In our example, we have three variables describing expectations, and three variables describing satisfaction.
Therefore, we could rename them to CUEX 1:3 and CUSA 1:3.
So now, in SEMinR, instead of writing this
c("blablabla")
we can write this
multi_items("CUEX", 1:3)
This might seem tedious in the beginning, but especially if you modify your model this will save you a lot of time.
---
