---
title: "What is Structural Equation Modeling"
subtitle: "Basic Introduction to SEM-in-R"
author: "André Calero Valdez and Lilian Kojan"
#institute: "RStudio, PBC"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    #lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
library(tidyverse)
library(seminr)
```


# What is Structural Equation Modeling?
1. Properties
2. Components
3. Process of estimating a model



???



# What are the advantages of SEM?

Combines

* factor analysis
* regression equation

Allows for

* mediation analysis
* causal interpretation

???


---
# Types of SEM

* CB-SEM
* VB-SEM / PLS-SEM


---
# Components of a SEM

* latent variables (constructs)
* manifest variables (indicators)

--
Relationships between those components form
* structural model: relationships between constructs (unidirectional in PLS-SEM)
* measurement model: relationships between constructs and their indicators


---
# Graphical representation of SEM - example with mobi data set


---

# Underlying math

---
# Process of estimating a SEM - sequence of our videos

---

---

# SeminR Model

```{r include=FALSE}
set.seed(123)
  mobi <- mobi

  #seminr syntax for creating measurement model
  mobi_mm <- constructs(
    reflective("Satisfaction", multi_items("CUSA", 1:3)),
    composite("Complaints",   single_item("CUSCO")),
    reflective("Loyalty",      multi_items("CUSL", 1:3))
  )
  #seminr syntax for creating structural model
  mobi_sm <- relationships(
    paths(from = "Satisfaction", to = c("Complaints", "Loyalty")),
    paths(from = "Complaints",   to = "Loyalty")
  )

  mobi_pls <- estimate_pls(data = mobi,
                           measurement_model = mobi_mm,
                           structural_model = mobi_sm)

  mobi_boot <- bootstrap_model(mobi_pls, nboot = 100, cores = 1)
```

---

# Review
