---
title: "SEM-in-R Basics"
date: "`r Sys.Date()`"
#output:
#  rmdformats::downcute:
#    self_contained: true
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print = "75")
opts_chunk$set(echo = FALSE,
	             cache = TRUE,
               prompt = FALSE,
               tidy = TRUE,
               comment = NA,
               message = FALSE,
               warning = FALSE)
opts_knit$set(width = 75)
```


# How this course works
This course is designed to help you learn structural equation modeling (SEM) using SEMinR.
Since it makes sense to learn things in small chunks. We split the things you need to learn into
smaller pieces. Each will be taught in a short video. 

The course is organized along the steps that you would usually take when using SEM in a research problem or in an applied setting. However, it makes sense first get a basic understanding of the overall process before diving into the details.
Therefore, some material is listed as "advanced material". The course is designed in a way
that you do not need to know advanced materials in a first pass.
It does make sense to skip advanced materials in your first iteration. Then when you feel comfortable with the basics, feel free to return for advanced materials.

We organized it this way, as this organization should help you come back for reference later.
Nobody can remember everything, so this should also serve as an easy documentation when you return to SEM after having been busy with other things for a while. We've been there. 

Expert materials are designed to support you after you have collected some experience in structural equation modeling. 






# What is SEM?


This video is just to demo how content could be included. Ignore it :)

<iframe width="640" height="360" src="https://www.youtube.com/embed/5ZXsIRI4Wlc" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen data-external="1"></iframe>


```{r}

```


# Measurementmodel and structural model

# Advanced Materials

## Causal relationships

## CBSEM vs VBSEM

## Exogeneous vs endogeneous variables

